Pyrovelocity PyPI user notebook template

### 2024-03-20 scripts/examples/pypi_template/pyrovelocity_pypi_user_template.qmd
feat(pypi_template): generate and display figure outputs Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

---
title: Pyrovelocity PyPI user notebook template
toc: true
number-sections: true
highlight-style: pygments
csl: bibstyle.csl
lightbox: auto
format:
  nature-pdf:
    natbibstyle: sn-nature
    # classoption:
    #   - lineno
    cite-method: citeproc
    keep-tex: true
  html:
    mainfont: Latin Modern Roman
    code-fold: false
    html-math-method: katex
  docx: default
execute: 
  eval: true
  warning: false
  error: false
  cache: true
  keep-ipynb: true
author:
  - name: Pyrovelocity Team
abstract: |
  This notebook demonstrates how to run pyrovelocity in a Jupyter notebook.
  It attempts to support both Google Colab and local or remote Jupyter kernel servers.
keywords: [single-cell genomics, probabilistic modeling, dynamical systems]
bibliography: references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

@@

```{python}
#| code-fold: true
import importlib.util
import subprocess
import sys

@@

```{python}
#| code-fold: true
import os

@@

```{python}
#| output: false

@@

```{python}
#| code-fold: true
#| output: false

@@

```{python}
help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
```

Before we start, we will set environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples.

```{python}
import os
os.environ["PYROVELOCITY_TESTING_FLAG"] = "True"
```

@@

```{python}
from importlib import reload
import yaml

import pyrovelocity.utils
reload(pyrovelocity.utils)
from pyrovelocity.utils import (
  print_config_tree,
  print_docstring,
)

import pyrovelocity.workflows.main_workflow
reload(pyrovelocity.workflows.main_workflow)
from pyrovelocity.workflows.main_workflow import (
  download_data,
  preprocess_data,
  train_model,
  postprocess_data,
  summarize_data,
)
```

@@

```{python}
from pyrovelocity.workflows.main_configuration import pancreas_configuration
```

@@

```{python}
pancreas_configuration_dict = pancreas_configuration.to_dict()
print_config_tree(pancreas_configuration_dict, "pancreas configuration")

@@

```{python}
print_config_tree(pancreas_configuration.download_dataset.to_dict(), "download dataset")
```

@@

```{python}
data = download_data(download_dataset_args=pancreas_configuration.download_dataset)
```

@@

```{python}
print_config_tree(pancreas_configuration.preprocess_data.to_dict(), "preprocess data")
```

@@

```{python}
print_docstring(pyrovelocity.preprocess.preprocess_dataset)
```

@@

```{python}
processed_data = preprocess_data(
  data=data,
  preprocess_data_args=pancreas_configuration.preprocess_data,
)
```

@@

```{python}
print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
```

@@

```{python}
print_docstring(pyrovelocity.train.train_dataset)
```

@@

```{python}
model_output = train_model(
  processed_data=processed_data,
  train_model_configuration=pancreas_configuration.training_configuration_2,
)
```

@@

```{python}
print_config_tree(model_output.to_dict(), "model output")
```

@@

```{python}
print_config_tree(pancreas_configuration.postprocess_configuration.to_dict(), "postprocess data")
```

@@

```{python}
print_docstring(pyrovelocity.postprocess.postprocess_dataset)
```

@@

```{python}
postprocessing_outputs = postprocess_data(
  preprocess_data_args=pancreas_configuration.preprocess_data,
  training_outputs=model_output,
  postprocess_configuration=pancreas_configuration.postprocess_configuration,
)
```

@@

```{python}
#| output: false
dataset_summary = summarize_data(
  preprocess_data_args=pancreas_configuration.preprocess_data,
  postprocessing_outputs=postprocessing_outputs,
  training_outputs=model_output,
)
```

@@

```{python}
print_config_tree(dataset_summary.to_dict(), "dataset summary")
```

### Vector fields

@@

```{python}
#| label: fig-vector-fields-uncertainty
#| echo: false
#| fig-cap: "Vector fields with uncertainty"

from IPython.display import Image, display

display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
```

A standard vector field plot with cluster annotations is shown in @fig-vector-field.

```{python}
#| label: fig-vector-field
#| echo: false
#| fig-cap: "Vector field"

display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
```

>>>>

---
title: Pyrovelocity PyPI user notebook template
toc: true
number-sections: true
highlight-style: pygments
csl: bibstyle.csl
lightbox: auto
format:
  nature-pdf:
    natbibstyle: sn-nature
    # classoption:
    #   - lineno
    cite-method: citeproc
    keep-tex: true
  html:
    mainfont: Latin Modern Sans
    code-fold: false
    html-math-method: katex
    embed-resources: true
  docx: default
  ipynb: default
format-links: false
execute: 
  eval: true
  warning: false
  error: false
  cache: true
  keep-ipynb: true
  keep-md: true
author:
  - name: Pyrovelocity Team
    email: team@pyrovelocity.net
abstract: |
  This notebook demonstrates how to run pyrovelocity in a Jupyter notebook.
  It attempts to support both Google Colab and local or remote Jupyter kernel servers.
keywords: [single-cell transcriptomics, probabilistic modeling, dynamical systems, RNA velocity]
bibliography: references.bib
jupyter:
  jupytext:
    cell_metadata_filter: all
    cell_metadata_json: true
    notebook_metadata_filter: all
    text_representation:
      extension: .qmd
      format_name: quarto
      format_version: 1.0
      jupytext_version: 1.16.0
  kernelspec:
    display_name: Python 3
    language: python
    name: python3
  language_info:
    name: python
  rise:
    scroll: true
    theme: black
  toc-autonumbering: true
  toc-showcode: false
  toc-showmarkdowntxt: false
---

@@

```{python}
#| label: install-package-helpers
#| code-fold: true
import importlib.util
import subprocess
import sys

@@

```{python}
#| label: install-pyrovelocity-colab
#| code-fold: true
import os

@@

```{python}
#| label: check-pyrovelocity-installation
#| output: false

@@

```{python}
#| label: check-pyro-installation
#| code-fold: true
#| output: false

@@

```{python}
#| label: print-pyrovelocity-docstring
help(pyrovelocity) # ?pyrovelocity # to open in side panel tab for reference
```

Before we start, we will adjust a few environment variables to ensure our first execution occurs in a lightweight test mode with a subset of observations, variables, training epochs, and posterior samples. We will also set a few log levels. Feel free to adjust these to `ERROR`, `DEBUG`, etc as needed.

```{python}
#| label: set-pyrovelocity-testing-flag
import os
os.environ["PYROVELOCITY_TESTING_FLAG"] = "True"
os.environ["PYROVELOCITY_LOG_LEVEL"] = "INFO"

import logging
logging.getLogger("jax").setLevel(logging.ERROR)
logging.getLogger("root").setLevel(logging.ERROR)
```

@@

```{python}
#| label: import-notebook-dependencies
from importlib import reload

import yaml
from IPython.display import Image
# from IPython.display import display

import pyrovelocity.utils
import pyrovelocity.workflows.main_workflow
reload(pyrovelocity.utils)
reload(pyrovelocity.workflows.main_workflow)

from pyrovelocity.utils import print_config_tree
from pyrovelocity.utils import print_docstring
from pyrovelocity.workflows.main_workflow import download_data
from pyrovelocity.workflows.main_workflow import postprocess_data
from pyrovelocity.workflows.main_workflow import preprocess_data
from pyrovelocity.workflows.main_workflow import summarize_data
from pyrovelocity.workflows.main_workflow import train_model
```

@@

```{python}
#| label: import-pancreas-configuration
from pyrovelocity.workflows.main_configuration import pancreas_configuration
```

@@

```{python}
#| label: print-and-save-pancreas-configuration
pancreas_configuration_dict = pancreas_configuration.to_dict()
print_config_tree(pancreas_configuration_dict, "pancreas configuration")

@@

```{python}
#| label: print-download-dataset-configuration
print_config_tree(pancreas_configuration.download_dataset.to_dict(), "download dataset")
```

@@

```{python}
#| label: execute-download-data-task
data = download_data(download_dataset_args=pancreas_configuration.download_dataset)
```

@@

```{python}
#| label: print-preprocess-data-configuration
print_config_tree(pancreas_configuration.preprocess_data.to_dict(), "preprocess data")
```

@@

```{python}
#| label: print-preprocess-data-docstring
print_docstring(pyrovelocity.preprocess.preprocess_dataset)
```

@@

```{python}
#| label: execute-preprocess-data-task
processed_data = preprocess_data(
  data=data,
  preprocess_data_args=pancreas_configuration.preprocess_data,
)
```

@@

```{python}
#| label: print-train-model-configuration
print_config_tree(pancreas_configuration.training_configuration_2.to_dict(), "train model 2")
```

@@

```{python}
#| label: print-train-model-docstring
print_docstring(pyrovelocity.train.train_dataset)
```

@@

```{python}
#| label: execute-train-model-task
model_output = train_model(
  processed_data=processed_data,
  train_model_configuration=pancreas_configuration.training_configuration_2,
)
```

@@

```{python}
#| label: print-model-output
print_config_tree(model_output.to_dict(), "model output")
```

@@

```{python}
#| label: print-postprocess-data-configuration
print_config_tree(pancreas_configuration.postprocess_configuration.to_dict(), "postprocess data")
```

@@

```{python}
#| label: print-postprocess-data-docstring
print_docstring(pyrovelocity.postprocess.postprocess_dataset)
```

@@

```{python}
#| label: execute-postprocess-data-task
#| output: false 
from pyrovelocity.utils import SuppressOutput

with SuppressOutput():
  postprocessing_outputs = postprocess_data(
    preprocess_data_args=pancreas_configuration.preprocess_data,
    training_outputs=model_output,
    postprocess_configuration=pancreas_configuration.postprocess_configuration,
  )
```

@@

```{python}
#| label: execute-summarize-data-task
#| output: false
dataset_summary = summarize_data(
  preprocess_data_args=pancreas_configuration.preprocess_data,
  postprocessing_outputs=postprocessing_outputs,
  training_outputs=model_output,
)
```

@@

```{python}
#| label: print-dataset-summary
print_config_tree(dataset_summary.to_dict(), "dataset summary")
```

### Data set review 
```{python}
#| label: import-ipython-display
#| echo: false
from IPython.display import Image, display
```

The maximum values of the unspliced and spliced count data for each gene across cells is shown in @fig-raw-counts.

<!-- ![Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms.](data/processed/pancreas_thresh_histogram.pdf.png){#fig-raw-counts} -->
```{python}
#| label: fig-raw-counts
#| echo: false
#| fig-cap: "Maximum spliced and unsliced transcript counts data for each gene across all cells with marginal histograms."
display(Image(filename=f"data/processed/pancreas_thresh_histogram.pdf.png"))
```

### Cell state transformation vector field estimates

A vector field estimate of cell state transformation probability derived from RNA velocity estimates with cluster annotations is shown in @fig-vector-field.

<!-- ![Vector field](reports/pancreas_model2/vector_field.pdf.png){#fig-vector-field} -->
```{python}
#| label: fig-vector-field
#| echo: false
#| fig-cap: "Vector field" 
display(Image(filename=f"{dataset_summary.data_model_reports.path}/vector_field.pdf.png"))
```

#### Vector field uncertainty estimates

@@

<!-- ![Vector fields with uncertainty estimates](reports/pancreas_model2/fig2_part1_plot.pdf.png){#fig-vector-fields-uncertainty} -->
```{python}
#| label: fig-vector-fields-uncertainty
#| echo: false
#| fig-cap: "Vector fields with uncertainty estimates."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part1_plot.pdf.png"))
```

### Gene selection

A *volcano plot* adapted for probabilistic inference of RNA velocity is shown in @fig-volcano. This plot shows the mean absolute error versus shared time correlation for each gene.

<!-- ![RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene](reports/pancreas_model2/volcano.pdf.png){#fig-volcano} -->
```{python}
#| label: fig-volcano
#| echo: false
#| fig-cap: "RNA velocity volcano plot showing mean absolute error versus shared time correlation for each gene."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/volcano.pdf.png"))
```

#### Splice state phase portraits

@fig-rainbow shows the splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes.

<!-- ![Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes](reports/pancreas_model2/rainbow.pdf.png){#fig-rainbow} -->
```{python}
#| label: fig-rainbow
#| echo: false
#| fig-cap: "Splice state phase portraits, estimated expression levels over shared time, and shared time correlation for a subset of selected genes."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/rainbow.pdf.png"))
```

#### Summary

A summary of the gene selection results is shown in @fig-gene-selection-summary.

<!-- ![Summary of gene selection results](reports/pancreas_model2/fig2_part2_plot.pdf.png){#fig-gene-selection-summary} -->
```{python}
#| label: fig-gene-selection-summary
#| echo: false
#| fig-cap: "Summary of gene selection results."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/fig2_part2_plot.pdf.png"))
```

### Uncertainty estimation

#### Parameter estimates

Estimates for parameter uncertainties for a subset of genes are shown in @fig-parameter-uncertainty.

<!-- ![Centered parameter uncertainties for a subset of genes](reports/pancreas_model2/parameter_uncertainties.pdf.png){#fig-parameter-uncertainty} -->
```{python}
#| label: fig-parameter-uncertainty
#| echo: false
#| fig-cap: "Centered parameter uncertainties for a subset of genes."
display(Image(filename=f"{dataset_summary.data_model_reports.path}/parameter_uncertainties.pdf.png"))
```

<!-- 
TODO: save png copy of shared time plot
#### Shared time

@fig-shared-time-uncertainty shows the uncertainty in the shared time estimates for each gene.

![Uncertainty in the shared time estimates for each gene](reports/pancreas_model2/shared_time.pdf){#fig-shared-time-uncertainty} 
-->
